---
title: "Data Wrangling - Tempe Accident Analysis"
author: "Abishek Rajaswamy Kuppusamy"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    theme: flatly
    highlight: haddock
    toc: yes
    toc_float: yes
---

```{r include = FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10)     # Knitting options
library(dplyr)
library(pander) 
library( ggmap )    # grab map tiles
library( viridis )  # color pallette for maps
library( ggplot2 )  # fancy graphics
library( ggthemes )

```

<br>

# Source Data

The following analysis uses data comprised of traffic accidents in the city of 
Tempe, Arizona. They constitute 28,470 observations and 32 variables. You can 
find more details in the dataset documentation at the 
[**Tempe Open Data Portal**](https://data.tempe.gov/dataset/high-severity-traffic-crashes-1-08). 

```{r include = FALSE}

# READ IN DATA

url <- paste0("https://github.com/DS4PS/Data-Science-Class/blob",
              "/master/DATA/TempeTrafficAccidents.rds?raw=true")

dat <- readRDS(gzcon(url(url)))     # Method per instructions

```

<br>

## Preview

We can explore the first six observations in the following table using function `head()`.

```{r}

head(dat)

```

<br>
<br>

# Preprocessing



```{r include = FALSE}

date.vec <- strptime(dat$DateTime, 
                     format = "%m/%d/%y %H:%M")       # Create date fields

dat$hour   <- format(date.vec, format = "%H")
dat$month  <- format(date.vec, format = "%b")
dat$day    <- format(date.vec, format = "%a")
dat$day365 <- format(date.vec, format = "%j")
dat$week   <- format(date.vec, format = "%V")

dat$day <- factor(dat$day, 
                  levels = c("Mon",
                             "Tue",
                             "Wed",
                             "Thu",
                             "Fri",
                             "Sat",
                             "Sun"))                  # Order weekdays

dat$hour12 <- format(date.vec, 
                     format="%l %p")                  # Create 12-hour format

time.levels <- c("12 AM", paste(1:11, "AM"), 
                 "12 PM", paste(1:11, "PM"))

dat$hour12 <- factor(dat$hour12, 
                     labels = time.levels)            # Order time intervals

age.labels <- paste0("Age ", 
                     c(16,18,25,35,45,55,65,75), "-", 
                     c(18,25,35,45,55,65,75,100) )

dat$age <- cut(dat$Age_Drv1, 
               breaks = c(16,18,25,
                          35,45,55,
                          65,75,100), 
               labels = age.labels)                   # Discretize age ranges

```

<br>
<br>

# Part I: Summaries

The following provides brief exploratory summaries on key data points.

<br>

## 1. Mondays Accidents

**Question:** *How many accidents happened on Mondays?*


<br>

```{r}

sum(dat$day == "Mon", 
    na.rm = TRUE)

```

<br>

**Answer:** **4,094** accidents occurred on Mondays.

<br>

## 2. Monday Accidents (%)

**Question:** *What proportion of accidents each week occur on Monday??*

<br>

```{r}


mean(dat$day== "Mon", 
    na.rm = TRUE)

```

<br>

**Answer:** **14.3%** of all accidents occur on Mondays.

<br>

## 3. Harmful Monday Accidents (%)

**Question:** *What proportion of accidents on Mondays result in harm??*


<br>

```{r}

dat %>%
  filter(day=="Mon") %>%
  mutate(harm_acc = Totalinjuries > 0 | Totalfatalities > 0) %>% summarize(mean_harm = mean(harm_acc))


```

<br>

**Answer:** **29.84%** of all Monday accidents have at least one casualty.

<br>

## 4. Most Common Accidents

**Question:** *What is the most common accident type on Mondays?*

<br>

```{r}

dat %>%
  filter(day == "Mon") %>%
  count(Collisionmanner) %>%
  arrange(desc(n))

```

<br>

**Answer:** **Rear End** collisions are the most common accident types.

<br>

## 5. Differences in Accidents

**Question:** *Are there differences in the proportion of accidents that result in harm each day of the week?*

<br>

```{r}

dat %>%
  group_by(day) %>%
  mutate(harm_acc = Totalinjuries > 0 | Totalfatalities > 0) %>% summarize(mean_harm = mean(harm_acc))

```

<br>

**Answer:** Significantly more accidents occur **on Thursday**.

<br>

## 6. Accident Stats by Weekday

Creating a table that reports the following for each day of the week.

* Total accidents
* Total injuries
* Total fatalities
* Proportion of harmful accidents (i.e. casualties)

<br>

```{r}

dat %>%
  group_by(day) %>%
  mutate(harm_acc = Totalinjuries > 0 | Totalfatalities > 0) %>%
  summarize(n = n(),
            injuries = sum(Totalinjuries),
            fatalities = sum(Totalfatalities),
            harm.rate = mean(harm_acc))

```

<br>
<br>

# Part II: Age Groups

The following provides summaries of accidents by age groups.

<br>

## 1. Accidents by Hour & Age

Creating a table of counts of accidents by time of day (`hour12`) and age of driver (`age`).

**Question:** *Which age group has the largest number of accidents at 7 AM?*

<br>

```{r}

dat %>% filter(hour12 == "7 AM") %>%
  group_by(age) %>%  summarize(n = n()) 
  
```

<br>

**Answer:** Drivers of **Age 18-25** have the greatest number of accidents from 7:00 to 7:59 AM.

<br>

## 2. Accidents by Hour & Age (%)

Creating a new table of time of day and age group that reports the proportion of accidents at "7 AM" (`hour12`) for each age group (`age`). *The proportions within each age group should sum to one.*

<br>
<br>


```{r echo = FALSE}

data.frame(age = c("Age 16-18", "Age 18-25", "Age 25-35"),
           hour12 = c("7 AM", "7 AM", "7 AM"),
           n = c(77, 408, 371),
           n.age = c(1458, 8796, 5456),
           n.hour = c(1606, 1606, 1606),
           p = c(0.05, 0.25, 0.23),
           p.age = c(0.05, 0.05, 0.07),
           p.hour = c(0.05, 0.25, 0.23)) %>%
  pander()

```

<br>
<br>


```{r}

dat %>% 
  count(age, hour12) %>%
  group_by(age) %>%
  mutate(n.age = sum(n)) %>%
  ungroup() %>%
  group_by(hour12) %>%
  mutate(n.hour = sum(n)) %>%
  mutate(p = round(n / n.hour, 2)) %>%
  mutate(p.age = round(n / n.age, 2)) %>%
  mutate(p.hour = round(n / n.hour, 2))%>%
  filter(hour12 == "7 AM")

```

<br>
<br>

# Part III: Rates of Harm

The following reports the accidents, casualties, proportion, and average 
casualties per harmful accident.

<br>

## 1. Accidents by Hour

Visualizing total accidents by time of day (`hour`).


<br>

```{r}

dat %>%
  group_by(hour) %>%
  summarize(n = n()) %>%
  plot(type = "b",
       bty = "n",
       pch = 19,
       cex = 2,
       xlab = "Hour",
       ylab = "Total Accidents",
       main = "Total Accidents by Time of Day")

```

<br>

## 2. Total Casualties by Hour

Visualizing total injuries and fatalities by time of day (`hour`).

<br>

```{r}
dat %>%
  group_by(hour) %>%
  mutate(Harm = Totalinjuries > 0 | Totalfatalities > 0) %>%
  summarize(Harm_acd = sum(Harm)) %>%
  plot(type = "b",
       bty = "n",
       pch = 19,
       cex = 2,
       xlab = "Hour of the Day",
       ylab = "Total Number of Passengers Hurt or Killed",
       main = "Total Injuries or Fatalities by Hour of the day")
```

<br>

## 3. Accidents with Casualties (%)

Visualizing the proportion of harmful collisions our of all collisions by time of day (`hour`).

<br>

```{r}

dat %>%
  group_by(hour) %>%
  mutate(Harm = Totalinjuries > 0 | Totalfatalities > 0) %>%
  summarize(Harm_acd1 = mean(Harm)) %>%
  plot(type = "b",
       bty = "n",
       pch = 19,
       cex = 2,
       xlab = "Hour of the Day",
       ylab = "Proportion of Accidents Resulting in Harm",
       main = "Proportion of Crashes that Result in Injuries or Fatalities")
```

<br>

## 4. Avg. Harm by Harmful Accident

Visualizing average injuries or fatalities per harmful collision.

<br>

```{r}

dat %>%
  filter(Totalinjuries > 0 | Totalfatalities > 0 ) %>%
  mutate( total_harm = Totalinjuries + Totalfatalities ) %>%
  group_by(hour) %>%
  summarize( average_injuries = ( sum(total_harm) / n()) ) %>%
  plot(type = "b",
       bty = "n",
       pch = 19,
       cex = 2,
       xlab = "Hour of the Day",
       ylab = "Ave. Number of Passengers Hurt",
       main = "Average Injuries or Fatalities Per Harmful Crash")
```

<br>
