---
title: "Descriptives and Operators"
author: "Abishek Rajaswamy Kuppusamy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10, 
                      fig.height = 6)

```

<br>

# Data

The following report identifies specific tax parcels in Downtown Syracuse, New York (USA).

View the dataset documentation here: [Syracuse, NY Tax Parcel Data](https://ds4ps.github.io/Data-Science-Class/DATA/syr_parcels.html).

<br>

## Required Packages

<br>

```{r}

library(sp)
library(rgdal)
library(geojsonio)                            # Load required libraries

```

<br>

## Import Spatial Polygons

We read in spatial polygons (`"sp"`) with function `geojson_read()` to map 
tax parcel shapes in Syracuse's Downtown area.

<br>

```{r cache = TRUE}

url <- paste0("https://raw.githubusercontent.com/DS4PS/Data",
              "-Science-Class/master/DATA/downtown-syr.geojson")

downtown <- geojson_read(url, what = "sp")    # Load parcel data

plot(downtown,  
     border = "gray60",
     col = "gray80")                          # Visualize parcels

```

<br>
<br>

### Part 1: Tax Parcel Acres

Parcels with more than one acre are determined with variable `acres` and condition `> 1`.

* Each value in variable `acres` is tested as greater than one, or `> 1`
* Values in `acres` greater than one become `TRUE`, otherwise `FALSE`
* All `TRUE` and `FALSE` values are stored in object `these`

<br>

```{r}

result <- downtown$acres > 1                  # Each value in 'acres' is tested

```

<br>

All `TRUE` and `FALSE` values, for each parcel, are converted into colors. 

<br>

```{r}

group.colors <- ifelse(test = result,         # 'test =' accepts TRUE or FALSE values
                       yes = "firebrick",     # 'yes =' converts values that are TRUE
                       no = "gray80")         # 'no =' converts values that are FALSE

```

<br>

All tax parcels in `downtown` are mapped with `plot()`.  Each parcel has one of 
two colors per the above code.

<br>

```{r}

plot(downtown,                                # Plots object 'downtown'
     border = NA,
     col = group.colors)                      # Red when TRUE; gray when FALSE

```

<br>
<br>

### Part 2: Parcels with Single Families

Using variable `landuse` to determine how a tax parcel is used.

<br>

```{r}

unique(downtown$landuse)                      # All unique values in 'landuse'

```

<br>

One of the values in `landuse` is "Single Family".  Therefore, the test is 
whether or not `landuse` is exactly equal to "Single Family".

<br>

```{r}

result <- downtown$landuse == "Single Family" # Test if 'landuse' is "Single Family"

group.colors <- ifelse(test = result,         # Provide TRUE or FALSE test results
                       yes = "firebrick",     # If TRUE, make "firebrick"
                       no = "gray80")         # If FALSE, make "gray80"

plot(downtown,
     border = NA, 
     col = group.colors)                      # Plot with respective colors

```

<br>


## Part 3: Vacant Lots

**Question:** *Where are the majority of vacant lots located in the* 
*downtown?*

<br>

```{r}

unique(downtown$landuse)                      # Print unique values in 'landuse'

```

<br>

Now, we create our test statement, `== "Vacant Land"`, and store the results in `result`.

<br>

```{r}

result <- downtown$landuse == "Vacant Land"

group.colors <- ifelse(test = result, 
                       yes = "firebrick", 
                       no = "gray80")

plot(downtown,
     border = NA, 
     col = group.colors)

```

<br>
<br>

## Part 4: Parking Lots

**Question:** *How many parking lots are in downtown Syracuse?*

**Answer:** There are **78** parking lots in downtown Syracuse.

<br>

```{r}

# Use function: 'sum()'
# Use variable: 'landuse'


result <- downtown$landuse == "Parking"

Total_lots <- sum(result)

Total_lots

group.colors <- ifelse(test = result, 
                       yes = "firebrick", 
                       no = "gray80")

plot(downtown,
     border = NA, 
     col = group.colors)

```

<br>
<br>

## Part 5: New Construction

**Question:** *Where is new construction located in the city?*

<br>

```{r}

# Use variable: 'yearbuilt'
# Find the number of buildings constructed either after or during and after than 1980

result <- downtown$yearbuilt >= 1980

New_Buliding <- sum(result, na.rm = T)

New_Buliding

group.colors <- ifelse(test = result, 
                       yes = "firebrick", 
                       no = "gray80")

plot(downtown,
     border = NA, 
     col = group.colors)

```

<br>
<br>

**Part 6:** *What proportion of commercial properties are built since 1980?*

**Answer:** **7.812%** of commercial properties were built since 1980.

<br>

```{r}

# Use variable: 'landuse'
# Use variable: 'yearbuilt'
# Count only commercial properties with buildings constructed during or later than 1980
Commercial_Building <- downtown$yearbuilt >= 1980 & downtown$landuse == "Commercial"
mean(Commercial_Building, na.rm = T)
```

<br>
<br>

## Part 7: Parcels Valued Over $10M

**Question:** *How many parcels have assessed values over $10 million?*

**Answer:** **16** parcels have assessed values over $10 M.

<br>

```{r}

# Use function: 'sum()'
# Use variable: 'assessedval'

result <- downtown$assessedval > 10000000

Total_Parcels <- sum(result, na.rm = T)

Total_Parcels

group.colors <- ifelse(test = result, 
                       yes = "firebrick", 
                       no = "gray80")

plot(downtown,
     border = NA, 
     col = group.colors)

```

<br>
<br>

## Part 8: Total Tax Delinquent Parcels

**Question:** *How many properties are delinquent on taxes?*

**Answer:** **57** properties are delinquent on taxes.

<br>

```{r}

# Use function: 'sum()'
# Use variable: 'amtdelinqt'

result <- downtown$amtdelinqt != 0

Prop_delinquent <- sum(result, na.rm = T)

Prop_delinquent

group.colors <- ifelse(test = result, 
                       yes = "firebrick", 
                       no = "gray80")

plot(downtown,
     border = NA, 
     col = group.colors)

```

<br>
<br>

## Part 9: Tax Delinquent Parcels

**Question:** *What proportion of parcels have delinquent tax payments owed?*

**Answer:** **14.65%** of parcels have delinquent tax payments owed.

<br>

```{r}

# Use function: 'mean()'
# Use variable: 'amtdelinqt'
Res <- mean(result, na.rm = T)
Res

```

<br>
<br>

## Part 10: Tax Delinquent Commercial Properties

**Question I:** *What proportion of commercial properties are delinquent on taxes?*

**Question II:** *What proportion of delinquent tax bills are owed by commercial parcels?*

<br>

**Answer I:** **11.96%** of commercial properties are delinquent on taxes. 

**Answer II:** **86.95%** of delinquent tax dollars are owed by commercial parcels.

<br>

```{r}

# Use function: 'mean()'
# Use variable: 'amtdelinqt'
# Use variable: 'landuse'

# For the first sub division tax-delinquent is commercial properties over all commercial properties
Com_deli <- (sum(downtown$landuse == "Commercial" & downtown$amtdelinqt !=0, na.rm = TRUE)) / (sum(downtown$landuse == "Commercial", na.rm = TRUE))

print(Com_deli)



# THe second subdivision is the tax dollars owed by commercial properties (a subset) over all tax dollars owed
Result1 <- downtown[downtown$landuse == "Commercial", 22]

Res1 <- sum(Result1$amtdelinqt, na.rm = TRUE)


Overall_tax <- sum(downtown$amtdelinqt, na.rm = TRUE)

Answer <- Res1 / Overall_tax
print(Answer)

```

<br>
<br>

## Part 11: Tax Delinquent Parcels by Land Use

**Question:** *How many of each land use type are delinquent on taxes? Printing in a table.*

<br>

```{r}

# Use function: 'table()'
# Use variable: 'amtdelinqt'
# Use variable: 'landuse'
Tab <- table(downtown$landuse,downtown$amtdelinqt != 0)
Answer <- Tab[,2]
Answer
```

<br>
<br>

