---
title: "Functions and Vectors"
author: "Abishek Rajaswamy Kuppusamy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
---

```{r include = FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

```

<br>

# Source Data

The following report analyzes tax parcel data from Syracuse, New York (USA).

View the "Data Dictionary" here: 
[**Syracuse City Tax Parcel Data**](http://data.syrgov.net/datasets/1b346804e1364a5eb85ccb53302e3c91_0)

<br>
<br>

# Importing the Data

The following code imports the Syracuse, NY tax parcel data using a URL.

```{r cache = TRUE}

url <- paste0("https://raw.githubusercontent.com/DS4PS/Data",
              "-Science-Class/master/DATA/syr_parcels.csv")

dat <- read.csv(url, 
                strings = FALSE)

```

<br>
<br>

# Previewing the Data

There are several exploratory functions to better understand our new dataset.

We can inspect the first 5 rows of these data using function `head()`.

<br>

```{r}

head(dat, 5)  

```

<br>

### Listing All Variables

Functions `names()` or `colnames()` will print all variable names in a dataset.

<br>

```{r}

names(dat)                # List all variables with 'names()'

```

<br>

### Previewing Specific Variables

We can also inspect the values of a variable by extracting it with `$`.

The extracted variable is called a "vector".

<br>

```{r}

head(dat$owner, 10)       # Preview a variable, or "vector"

```

<br>

### Listing Unique Values

Function `unique()` helps us determine what values exist in a variable.

<br>

```{r}

unique(dat$land_use)      # Print all possible values with 'unique()'

```

<br>

### Examining Data Structure

Function `str()` provides an overview of total rows and columns (dimensions), 
variable classes, and a preview of values.

<br>

```{r}

str(object = dat,
    vec.len = 2)          # Examine data structure with 'str()'

```

<br>
<br>


## Part 1: Total Parcels

**Question:** *How many tax parcels are in Syracuse, NY?* 

**Answer:** There are **41,502** tax parcels in Syracuse, NY. 

<br>

```{r}

# exploratory functions: 'dim()', 'nrow()', or 'str()'
dim(dat)[1]

```

<br>

## Part 2: Total Acres

**Question:** *How many acres of land are in Syracuse, NY?*

**Answer:** There are **12,510.49** acres of land in Syracuse, NY.

<br>

```{r}

# Passing a numeric variable to function 'sum()', with argument 'na.rm = TRUE'
sum(dat$acres, na.rm = TRUE)
```

<br>

## Part 3: Vacant Buildings

**Question:** *How many vacant buildings are there in Syracuse, NY?*

**Answer:** There are **1888** vacant buildings in Syracuse, NY.

<br>

```{r}

# Passing a numeric variable to function 'sum()', with argument 'na.rm = TRUE'
sum(dat$vacantbuil, na.rm = TRUE)
```


## Part 4: Tax-Exempt Parcels

**Question:** *What proportion of parcels are tax-exempt?*

**Answer:** **10.7%** of parcels are tax-exempt.

<br>

```{r}

# Passing a logical ('TRUE' or 'FALSE') variable to function 'mean()', with argument 'na.rm = TRUE'
mean(dat$tax.exempt, na.rm = TRUE)
```

<br>

## Part 5: Neighborhoods & Parcels

**Question:** *Which neighborhood contains the most tax parcels?*

**Answer:** **Eastwood** contains the most tax parcels.

<br>

```{r}

# Passing the appropriate variable to function 'table()'
table(dat$neighborhood)
```

<br>

## Part 6: Neighborhoods & Vacant Lots

**Question:** *Which neighborhood contains the most vacant lots?*

**Answer:** **Near Westside** contains the most vacant lots.

<br>

```{r}

# Passing two variables to function 'table()', separated by a comma

table(dat$neighborhood, dat$land_use)

```

<br>
<br>

